A Computational Account of Default Mode Function\\ by Control Theory and Reinforcement Learning

### 2016-10-23 paper/paper.tex
restructured free-energy principles + misc passes on tex

\subsection{Demystifying the free-energy principle and active-inference}
The so-called free-energy principle in its present form (including notions like ``generative density'', ``recognition density'', etc.) can be traced back to works of Dayan \& Hinton \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial directional deep belief-net (DBN).

In this section, we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in building their Helmholtz machine. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For example the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial, e.g) wake-sleep algorithm to train the underlying Helmholtz machine, etc.

@@

The following critics can be made:
\begin{itemize}
  \item
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
  \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
  \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
    technical difficulty in getting derivatives of loss function w.r.t
    recognition weights $\W^R$).
  \item This difficulty was removed in the 2010s by Kingma et al., an other groups, via a ``reparametrization trick''.
    \end{itemize}

\subsection{Free-energy minimization via variational auto-encoders}
Following Kingma et al. 2014 \citep{kingma2013auto}, define the data-dependent auxiliary random function
\begin{eqnarray}
  f_{G,R}(., \x) :\z \mapsto \log(p_G(\z,\x)) - \log(p_R(\z|\x)).
\end{eqnarray}
Then we can rewrite the
variational free-energy as
\begin{eqnarray*}
  \begin{split}
    F_G^R(\x) &:= \langle E_G(., \x) \rangle_{p_R(. |\x)} - \mathcal H(p_R(. |\x)) = \langle E_G(., \x) + \log(p_R(.|\x))\rangle_{P_R(.|\x)}\\
    &=\langle -\log(p_G(.,\x)) + \log(p_R(.|\x))\rangle_{P_R(.|\x)}\\
    &= -\langle f_{G,R}\rangle_{P_R(.|\x)} \approx -\frac{1}{M}\sum_{m=1}^Mf_{G,R}(\z^{(m)}), \text{ with }\z^{(1)},\ldots,\z^{(M)} \sim p_R(.|\x), \text{ and }M \rightarrow \infty.
    \end{split}
\end{eqnarray*}

\begin{mdframed}
  \textbf{Problem:} How do we sample from the recognition density $p_R(.|\x)$ in such a way that the sampling process is differentiable w.r.t the weights of the recognition network $\W^R$ ?
\end{mdframed}
\paragraph{Solution: The reparametrization trick.}
\begin{itemize}
\item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
\item Set $\phi = g(\W^{R}, \x, \epsilon)$, where $g$ is an appropriate class $\mathcal C^1$ function
  \begin{itemize}
  \item results in a sample $\phi \sim p_R(.|\x)$, from the correct
    posterior
    \end{itemize}
\end{itemize}

@@

%% To conclude this section, let's note that \citep{braun2011path} show how their free-energy framework (on paths) links with the well-known Hamilton-Jacobi-Bellman optimal control framework. For example, one can re-derive the Linear Quadratic Gaussian (LQG) controller, which is a generalization of the LQR in \eqref{eq:hj}...

@@

\section{Relation to other statistical models (Elvis)}
\subsection{Control theory.}
\begin{itemize}
  \item The model proposed in \citep{betzel2016} can be cast in the form \eqref{eq:hj}, with the particular choice $\Q = \textbf{I} = \rho^{-1}\R$ and $\Q_{\text{fin}} = \textbf{0}$, but with a rather ad-hoc handle on stability issues and choice driver nodes...
\end{itemize}

\subsection{Reinforcement Learning.}
abc

\subsection{Free Energy Principle.}
\begin{itemize}
  \item Closely related to Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  \item A common point between the Free Energy principle and our framework
  is placing the minimization of suprise at the core of brain function.
  In RL, surprise minimization reduces to accurate prediction of
  rewarding outcomes in the future.
  \item Ortega et al. 2013 \citep{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \citep{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
\item It is difficult to compute the free energy principle
in practice, which makes it physiologically less plausible.
  \end{itemize}

>>>>

\section{Relation to other statistical models (Elvis)}
\subsection{Free-energy principles!}
The so-called free-energy principle in its present form (including notions like ``generative density'', ``recognition density'', etc.) can be traced back to works of Dayan \& Hinton \citep{dayan1995helmholtz} in which they introduced the so-called \textit{Helmholtz machine}, a hierarchical factorial directional deep belief-net (DBN).

In this subsection, we will develop from first-principles, the bare-bones minimalistic ideas needed to build a free-energy principle for general decision-making. This ideas were first developed by Hinton et al. in the early 90s in building their Helmholtz machine. Theories like Friston's free-energy principle and active-inference will then emerge as particular instances of this general framework, with particular design choises. For example the Friston theory axiomatizes that the brain uses a (problematic, as it implicitly assumes that posterior of each hidden unit is factorial, e.g) wake-sleep algorithm to train the underlying Helmholtz machine, etc.

@@

The following critics can be made:
\begin{itemize}
  \item
    As noted by Dayan et al. (\textit{Variants of Helmholtz machines}), the inter-neuronal intra-layer independence assumption which is at the center of the HM becomes severely problematic as it is agnostic to the known organization of cortical layers...
  \item A drawback of the wake-sleep algorithm is that it requires a concurrent models (generative and recognitiion), which together do not correspond to optimization of (a bound of) the marginal likelihood (because of the incorrect KL used therein, etc.).
  \item Also, note that the wake-sleep algorithm doesn't do backprop! This is due to
    technical difficulty in getting derivatives of loss function w.r.t
    recognition weights $\W^R$).
  \item This difficulty was removed in the 2010s by \citep{kingma2013auto}, an other groups, via a ``reparametrization trick''.
    \end{itemize}

\subsection{Free-energy minimization via variational auto-encoders}
\label{sec:vae}
Here, we present a possible way to aleviate some conceptual and computational issues with the
free-energy framework presented thus far, by using auto-encoder theory.
Following Kingma et al. 2014 \citep{kingma2013auto}, define the data-dependent auxiliary random
function
\begin{eqnarray}
  f_{G,R}(., \x) :\z \mapsto \log(p_G(\z,\x)) - \log(p_R(\z|\x)).
\end{eqnarray}
Then we can rewrite the
variational free-energy as
\begin{eqnarray*}
  \begin{split}
    F_G^R(\x) &:= \langle E_G(., \x) \rangle_{p_R(. |\x)} - \mathcal H(p_R(. |\x)) = \langle E_G(., \x) + \log(p_R(.|\x))\rangle_{P_R(.|\x)}\\
    &=\langle -\log(p_G(.,\x)) + \log(p_R(.|\x))\rangle_{P_R(.|\x)}\\
    &= -\langle f_{G,R}\rangle_{P_R(.|\x)} \approx -\frac{1}{M}\sum_{m=1}^Mf_{G,R}(\z^{(m)}), \text{ with }\z^{(1)},\ldots,\z^{(M)} \sim p_R(.|\x), \text{ and }M \rightarrow \infty.
    \end{split}
\end{eqnarray*}

\begin{mdframed}
  \textbf{Problem:} How do we sample from the recognition density $p_R(.|\x)$ in such a way that the sampling process is differentiable w.r.t the weights of the recognition network $\W^R$ ?
\end{mdframed}
\paragraph{Solution: The reparametrization trick.}
\begin{itemize}
\item Choose $\epsilon \sim p_{\text{noise}}$ (noise distribution, independent of $\W^R$!)
\item Set $\z = g(\W^{R}, \x, \epsilon)$, where $g$ is an appropriate class $\mathcal C^1$ function
  \begin{itemize}
  \item results in a sample $\z \sim p_R(.|\x)$, from the correct
    posterior
    \end{itemize}
\end{itemize}

@@

%% To conclude this section, let's note that \citep{braun2011path} show how their free-energy framework (on paths) links with the well-known Hamilton-Jacobi-Bellman optimal control framework. For example, one can re-derive the Linear Quadratic Gaussian (LQG) controller, which is a generalization of the LQR in \eqref{eq:hj}...
\subsubsection{Concluding remarks}
In summary:
\begin{itemize}
  \item Friston's free-energy principle: brain as inference engine
  biological systems, including brains, must minimize the long-term average of surprise;
  \item A common point between the Free Energy principle and our framework
  is placing the minimization of suprise at the core of brain function.
  In RL, surprise minimization reduces to accurate prediction of
  rewarding outcomes in the future.
  \item Ortega et al. 2013 \citep{ortega2013thermodynamics} ``Thermodynamics as a theory of decision-making with information-processing costs''
\item Friston's critic of RL \citep{fristonAIorRL} (he proposes "active inference''): "This equation (equation 18) comes from the theory of dynamic programming,
pioneered in the 1950s by Richard Bellman and colleagues [2]. To
optimise control a~pð Þ x~ under this formulation, we have to: (i)
assume the hidden states are available to the agent and (ii) solve
Equation 18 for the value-function. Solving for the value-function
is a non-trivial problem and usually involves backwards induction
or some approximation scheme like reinforcement learning [4–6].
The free-energy formulation circumvents these problems by
prescribing the policy in terms of free-energy, which encodes
optimal control (Equation 6)"
\item It is difficult to compute with the free energy principle
  in practice, which makes it physiologically less plausible. Auto-encoder theory presented
 \ref{sec:vae} can help to remove these barriers.
  \end{itemize}

\subsection{Control theory}
\begin{itemize}
  \item Cite stuff from Bert K.
  \item The model proposed in \citep{betzel2016} can be cast in the form \eqref{eq:hj}, with the particular choice $\Q = \textbf{I} = \rho^{-1}\R$ and $\Q_{\text{fin}} = \textbf{0}$, but with a rather ad-hoc handle on stability issues and choice driver nodes...
\end{itemize}

\subsection{Other Leinforcement Learning literature}
abc

@@
